---
title: "Debugging"
subtitle: "âš”<br/>in R"
author: "Julia Piaskowski"
date: "2022-03-08"
output:
  xaringan::moon_reader:
    lib_dir: libs
    mathjax: null
    nature:
      highlightStyle: github
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

# When you encounter an R error: 

1. Check package documentation
1. Ask around: RStudio community, R4DS slack, Stack Overflow, your local oracle
1. Check source code (maybe)

#### You will probably need to work with debugging functions eventually

---
class: center, middle, inverse

## debugging = removing code errors  
<br> 
# How do you debug?  

---
# Retrace your steps

1. Go through a process step-by-step and see where things went wrong
1. Start with the action that occurred that first and proceed to the second action, the third, etc...

---
# Retrace Example

```{r include=FALSE, echo=FALSE}
library(dplyr); library(tidyr); library(purrr)
```


[Agricultural Data from Tidy Tuesday](https://github.com/rfordatascience/tidytuesday/tree/master/data/2020/2020-09-01)
```{r, eval=FALSE, include=FALSE, echo=TRUE}
# install.packages"(tidytuesdayR")
agdata <- tidytuesdayR::tt_load('2020-09-01')
saveRDS(agdata, file = "data/agdata.rds")
```


```{r, error=TRUE}
library(dplyr); library(tidyr); library(purrr)
agdata <- readRDS(here::here("static", "slides", "data", "agdata.rds"))
yields <- agdata[[4]]
colnames(yields) <- gsub("\\(tonnes per hectare\\)", "", names(yields))

wheat <- yields %>% tidyr::pivot_wider(id_cols = Year, 
                                  names_from = Entity,
                                  values_from = Wheat) %>% 
  purrr::discard(~all(is.na(.))) %>% 
  mutate(ethiopia = case_when(
    is.na(Ethiopia) ~ `Ethiopia PDR`,
    TRUE ~ Ethiopia)) %>% select(-Ethiopia, -`Ethiopia PDR`)
```

What went wrong? 

---
# Fixed:

```{r, highlight.output=2}
yields <- agdata[[4]]
colnames(yields) <- gsub(" \\(tonnes per hectare\\)", "", names(yields))

wheat <- yields %>% tidyr::pivot_wider(id_cols = Year, 
                                  names_from = Entity,
                                  values_from = Wheat) %>% 
  purrr::discard(~all(is.na(.))) %>% 
  mutate(ethiopia = case_when(
    is.na(Ethiopia) ~ `Ethiopia PDR`,
    TRUE ~ Ethiopia)) %>% select(-Ethiopia, -`Ethiopia PDR`)
```

---
# Another retrace example

**Working with nested functions makes debugging hard**

#### Hard: 
```{r eval=FALSE}
result <- function4(function3(function2(function1(my_object, extra_args = 1),more_args = TRUE), even_more_args = "show_me_the_money"), final_arg = FALSE)
```
--
#### Easier
```{r eval=FALSE}
pre_result1 <- function1(my_object, extra_args = 1)
pre_result2 <- function2(pre_result1, more_args = TRUE)
pre_result3 <- function3(pre_result2, even_more_args = "show_me_the_money")
result <- function4(pre_result3, final_arg = FALSE)
```
*(avoid writing nested functions)*  

---
# Consider writing a function if....

* You need to do something repeatedly
* that thing that needs to be repeated has several steps
* If you have to *cut-paste-replace* more than twice, consider a function to automate that process


---
# Why write functions? 

* To save yourself work  
* to avoid mistakes  
  
ðŸ‘  DRY = **D**on't **R**epeat **Y**ourself  
<br>
  
ðŸ‘Ž  WET = **W**rite **E**verything **T**wice / **We** **E**njoy **T**yping  

.footnote[[more instructions](https://agstats.io/post/writing-r-functions/) on writing functions]

---
# Crash course in functions

R functions follow a general structure:

```{r, eval=FALSE, out.width='75%'}
my_function_name <- function(arg1, arg2) {
  final_output <- action(arg1, arg2)
  return(final_output)
}
```

Their usage:
```{r, eval = FALSE}
my_result <- my_function_name(arg1 = ..., arg2 = ...)
```

---
# More intro to functions

It helps to describe the arguments (but, first we need an actual function to help us define the argument)
```{r}
my_function_name <- function(arg1, arg2) {
  # arg1 & arg2 = numeric vectors of identical lengths
  plot(arg1, arg2)
}
```

*(usually these functions are slightly more complicated)* 
 
```{r, out.height="60%"}
my_function_name(1:10, 1:10 + rnorm(10))
```


---
class: center, middle,  inverse

# [exercise](exercises/exercise-4.html)

---
## What to do when function errors happen

* `traceback`

```{r, error=TRUE}
f <- function(x) x + 1
g <- function(x) f(x)^3
g("a")

#traceback()
```


---
class: center, middle, inverse


# [mini-exercise](exercises/exercise-5.html)


---
# Trace

* sometimes you only want R to alert you when something specific happens
* this examines `print()` when it receives numeric input that is greater than or equal to 3. 

```{r}
trace(print, quote(if (is.numeric(x) && x >= 3) 
  cat("We have a problem: \n")), print = FALSE)
```

```{r}
print(4)
```

Remember to turn `trace` off: 
```{r,}
untrace(print)
```

---
## What to do when function errors happen

* sprinkle `print()` and/or `str()` statements in your function
(not a formal debug method, just something that works)

```{r}
log_hist <- function(var, color = "darkcyan", do_log=TRUE) {
  if(do_log == TRUE) {
    print("doing the log var")
    logvar = log(var)
    str(logvar)
    hist(logvar, col = color); abline(v = mean(logvar), lwd = 2, col = "red3")
  }
  else hist(var, col = color); abline(v = mean(var), lwd = 2, col = "red3")
}
```

--

```{r, error=TRUE}
log_hist(rnorm(100), do_log=TRUE)
```

---
## `browse()`

A more powerful way to examine objects in a function:

```{r}
fertil <- agdata[[2]]
colnames(fertil) <- c("country", "code", "year", 
                      "cereal_yield", "N_appl")
```

```{r}
my_lm_fun <- function(df, country){
  m1 = lm(country ~ year, data = df)
  a1 = anova(m1)
  r2 = summary(m1)$r2
  return(list(coefficients = coef(m1), anova = a1, rsq = r2))
}
```

```{r, error=TRUE}
my_lm_fun(df = fertil, country = "Canada")
```

---
class: center, middle, inverse

# [exercise](exercises/exercise-6.html)

---
## How do you examine an error for a function you didn't write? 

* `debug()` 

```{r, eval=FALSE}
debug(stats::lm)
```

```{r, eval=FALSE}
test <- lm(cereal_yield ~ N_appl, data = fertil)
```

Turn if off: 
```{r, eval=FALSE}
undebug(stats::lm)
```

---
# You can put this in your .Rprofile!

e.g. 
```{r, eval=FALSE}
options(error = traceback)
```
or
```{r, eval=FALSE}
options(error = browser)
```

....and more


---
class: center, middle, inverse

<iframe src="https://giphy.com/embed/ReImZejkBnqYU" width="530" height="300" frameBorder="0" class="giphy-embed" allowFullScreen></iframe><p><a href="https://giphy.com/gifs/angry-computer-frustrated-ReImZejkBnqYU">image credit: GIPHY</a></p>

## Keep Calm and Debug On

---
# Miscellaneous R tip: running R scripts in a terminal

```{r, eval = FALSE}
Rscript path/to/my/script.R 
```

```{r, eval=FALSE}
Rscript -e "Sys.Date()"
```


