---
title: "Debugging"
subtitle: "⚔<br/>in R"
author: "Julia Piaskowski"
date: "2022-03-08"
output:
  xaringan::moon_reader:
    lib_dir: libs
    mathjax: null
    nature:
      highlightStyle: github
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

# When you encounter an R error: 

1. check package documentation
1. ask around: RStudio community, R4DS slack, Stack Overflow, your local oracle
1. check source code

##### but, eventually, you will probably need to work with debugging functions

---
class: center, middle, inverse

## debugging = removing code errors  
<br> 
# How do you debug?  

---
# Retrace your steps

1. Go through a process step-by-step and see where things went wrong
1. Start with the action that occurred that first and proceed to the second action, the third, etc...

---
# Retrace Example

[Agricultural Data from Tidy Tuesday](https://github.com/rfordatascience/tidytuesday/tree/master/data/2020/2020-09-01)
```{r eval=FALSE}
# install.packages"(tidytuesdayR")
library(dplyr); library(tidyr); library(purrr)
agdata <- tidytuesdayR::tt_load('2020-09-01')
yields <- agdata[[4]]
colnames(yields) <- gsub("\\(tonnes per hectare\\)", "", names(yields))

wheat <- yields %>% tidyr::pivot_wider(id_cols = Year, 
                                  names_from = Entity,
                                  values_from = Wheat) %>% 
  purrr::discard(~all(is.na(.))) %>% 
  mutate(ethiopia = case_when(
    is.na(Ethiopia) ~ `Ethiopia PDR`,
    TRUE ~ Ethiopia)) %>% select(-Ethiopia, -`Ethiopia PDR`)
```
---
# Another retrace example

#### Hard: 
```{r eval=FALSE}
result <- function4(function3(function2(function1(my_object, extra_args = 1), more_args = TRUE), even_more_args = "show_me_the_money"), final_arg = FALSE)
```
--
#### Easier
```{r eval=FALSE}
pre_result1 <- function1(my_object, extra_args = 1)
pre_result2 <- function2(pre_result1, more_args = TRUE)
pre_result3 <- function3(pre_result2, even_more_args = "show_me_the_money")
result <- function4(pre_result3, final_arg = FALSE)
```
*(avoid writing nested functions)*  

---
# Consider writing a function if....

* You need to do something repeatedly
* that thing that needs to be repeated has several steps
* If you have to cut-paste-replace more than twice, consider a function to automate that process


---
# Why write functions? 

* To save yourself work  
* to avoid mistakes  
  
:thumbs-up: DRY = **D**on't **R**epeat **Y**ourself  
<br>
  
:thumbs-down: WET = **W**rite **E**verything **T**wice / **We** **E**njoy **T**yping  

---
# Crash course in functions

R functions follow a general structure:

```{r, eval=FALSE}
my_function_name <- function(arg1, arg2) {
  final_output <- action(arg1, arg2)
  return(final_output)
}
```

Their usage:
```{r, eval = FALSE}
my_result <- my_function_name(arg1 = ..., arg2 = ...)
```

---
# More intro to functions

It helps to describe the arguments (but, first we need an actual function to help us define the argument)
```{r}
my_function_name <- function(arg1, arg2) {
  # arg1 & arg2 = numeric vectors of identical lengths
  plot(arg1, arg2)
}
```

```{r}
plot(1:10, 1:10 + rnorm(10))
```
 *(usually these functions are slightly more complicated)* 

---
class: middle, center, inverse

# [exercise](exercises/exercise-4.html)

---





---
# Some future example

```{r, eval=FALSE}
fertil <- agdata[[2]]
colnames(fertil) <- c("country", "code", "year", 
                      "cereal_yield", "N_appl")
anova
```

---
class: center, middle, inverse

# Miscellaneous R tip: running R scripts in a terminal
```{r, eval = FALSE}
Rscript path/to/my/script.R 
```

```{bash}
Rscript -e "Sys.Date()"
```




